---
title: "ferols: A Robust M-Estimator with Huber Loss for Fixed-Effect Models"
format: gfm
---

```{r Setup, echo=FALSE}
library(kableExtra)
bmark_stats <- readRDS("utils/bmark_stats.rds")
bmark_timing <- readRDS("utils/bmark_timing.rds")
```

The R package `ferols` implements a robust M-Estimator for linear models with 
high-dimensional fixed effects, a Huber loss function and iteratively 
reweighted least squares (IRLS). It is inspired by and co-developed with the  
the [Stata package `robhdfe` by David Veenman](https://github.com/dveenman/robhdfe). 
Its user interface is designed to integrate tightly with the
[`fixest`](https://lrberge.github.io/fixest/) ecosystem.

It builds on our recent work

> Joachim Gassen and David Veenman (2026): Estimation Precision and 
> Robust Inference in Archival Research, SSRN Working Paper,
> http://dx.doi.org/10.2139/ssrn.4975569.

and combines robust M-estimation (Huber loss) with:

- high-dimensional fixed-effect absorption,
- fast estimation via `fixest::feols`,
- and a `fixest` style variance–covariance interface allowing for
  clustered sandwich standard errors.

For the first-step quantile regression that is used to obtain the scale 
estimate, `ferols()` implements the MM-QR estimation from 
[Machado and Santos Silva(2019)](https://www.sciencedirect.com/science/article/pii/S0304407619300648)
to combine quantile estimation with fixed effects, which 
[Rios-Avila, Siles, and Canavire-Bacarreza (2024)](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=4944894)
show can be extended to the multidimensional fixed effects setting. 

`ferols()` drops singleton observations by default. This implementation choice 
affects not only the degrees-of-freedom adjustment in the standard error 
calculation (see [Correia 2015](https://scorreia.com/research/singletons.pdf)), 
but in this case also ensures the scale parameter is correctly estimated with 
the MM-QR estimation. Singleton observations produce regression residuals that 
are equal to (or very close to) zero, causing the scale parameter to be 
understated when singleton observations are retained.


## Why?

Standard fixed-effects estimators are sensitive to outliers and heavy-tailed
error distributions. While `fixest` provides fast and reliable estimation
for large linear models with fixed effects, it does not currently offer
robust estimators.

The `ferols` package is designed to fill this gap. It is intended to be 
useful for settings with high-dimensional fixed effects (e.g. panel data with 
unit and time effects) and offers robust inference via Huber ψ/φ sandwich 
variance estimators, including clustered standard errors.


## Disclaimer 

`ferols` is not meant to replace 
[established R packages for robust regression](https://cran.r-project.org/web/views/Robust.html)
as these provide much more flexible, tested, and rigorous implementations. 
Rather it aims to fill a narrow gap as these packages currently do not allow
to absorb fixed effects during estimation and might thus run in 
performance/memory problems in settings with many fixed effects.

This package is under active development. The API, implementation details, and 
numerical behavior may and probably will change. Use with care.


## Installation (development version)

`ferols` is not on CRAN yet. To install the current development version from 
GitHub:

```{r, eval=FALSE}
install.packages("remotes")
remotes::install_github("joachim-gassen/ferols")
```


## Function to generate test panel data

For convenience, the package ships with the function `generate_panel_data()`. 
It generates panel data with a composite error term that has high kurtosis 
by default. The modeled data generating process is as follows

$$
y_{it} = \alpha_i + \lambda_t + \beta x_{it} + \gamma z_{it} + e_{it}
$$

where $\alpha_i$ and $\lambda_t$ are unit and time fixed effects, respectively. 
By default, $\beta = 1$ and $\gamma = 0$. The composite error is constructed as

$$
e_{it} = u_{it} + r_{it} + \eta_{it}
$$

with three components:

- **Unit component** ($u_{it}$): follows an AR(1) process over time within each 
  unit (serial correlation within units)

- **Time component** ($r_{it}$): follows an AR(1) process over units within each
  time period (cross-sectional dependence within time)

- **Idiosyncratic component** ($\eta_{it}$): i.i.d. noise

This construction induces two-way dependence consistent with the simulation 
logic in Petersen (2009).

To generate heavy-tailed errors, a share $p$ of observations is randomly 
selected. The composite errors $e_{it}$ of these observations are multiplied 
by a factor $m$. This produces a mixture error distribution with outliers.


## Example code

```{r exp1, fig.align='center'}
library(ferols)
library(moments)
df <- generate_panel_data(seed = 42, e_mult = 10)
cat(sprintf(
  "True beta: 1, true gamma: 0, kurtosis of composite error: %.2f", 
  kurtosis(df$e)
))
plot(
  df$x, df$y, xlab = "x", ylab = "y", 
  pch = 16, cex = 0.4, col = rgb(0, 0, 0, 0.2)
)
ferols(y ~ x + z | i + t, data = df, vcov = ~ i)
# For comparison: OLS has less precise estimates and larger standard errors
fixest::feols(y ~ x + z | i + t, data = df, vcov = ~ i)
```

## A comparison of the precision of OLS and robust M by increasing kurtosis of e

```{r exp2, fig.align="center"}
library(ferols)
library(fixest)
library(moments)

multiplers <- 1:10
res <- matrix(NA_real_, nrow = length(multiplers), ncol =  4)
colnames(res) <- c("m", "k_e", "se_x_feols", "se_x_ferols")
res[,1] <- multiplers
for (r in 1:nrow(res)) {
  df <- generate_panel_data(seed = 1234, e_mult = res[r, "m"])
  res[r, "k_e"] <- kurtosis(df$e)
  fit_ols <- feols(y ~ x + z | i + t, vcov = ~i, data = df)
  res[r, "se_x_feols"] <- fit_ols$se[1]
  fit_rols <- ferols(y ~ x + z | i + t, vcov = ~i, data = df)
  res[r, "se_x_ferols"] <- fit_rols$se[1]
} 
plot(
  main = "Standard error of x coefficient by kurtosis of composite error term",
  res[, "k_e"], res[, "se_x_feols"], type = "l", 
  xlab = bquote("kurtosis("~ e[i*t] ~ ")"), ylab = expression(SE(hat(beta))), 
  lwd  = 2, 
)
lines(
  res[, "k_e"], res[, "se_x_ferols"],
  lwd = 2, lty = 2, col = "blue"
)
points(res[, "k_e"], res[, "se_x_feols"], pch = 16)
points(res[, "k_e"], res[, "se_x_ferols"], col = "blue", pch = 16)
legend(
  "topleft",
  legend = c("OLS (feols)", "Robust M Huber (ferols)"),
  lty = c(1, 2), col = c("black", "blue"), lwd = 2, bty = "n"
)
```

## Benchmarking `ferols` with other R and Stata packages for Robust M-Estimation

`ferols` produces coefficient, standard error, and scale estimates that are 
identical within numerical precision to the ones created by packages that do not
aborb fixed effects (`MASS::rlm()`)^[To mimic the algorithm used by 
`MASS::rlm()`, `ferols` is called with the parameters `k = 1.345, 
scale_est = "ols", adj_rlm = TRUE, scale_update = TRUE, tol = 1e-4, 
vcov = "iid"`.], that absorb one dimension of fixed effects 
(Stata's `robreg`) and that absorb high-dimensional fixed effects (Stata's
`robhdfe`). Summary statistics of the relative differences of `ferols`'s 
estimates to these alternative packages ($rd_{est,package}$)

$$
rd_{est,package}  = \frac{est_{ferols} - est_{package}}{est_{package}}
$$

based on `{r} unique(bmark_stats$runs)` runs on independent random samples 
created by `generate_panel_data()`^[The code to create the benchmark data is at 
`utils/compare_ests.R`.] are presented in the table below.

```{r TabRelDiff, echo=FALSE}
df <- bmark_stats[1:9,c(1:2, 4:6)]
df[df[,2] == "est", 2] <- "coef"
df[, 3] <- sprintf("%.1e", df[, 3])
df[, 4] <- sprintf("%.1e", df[, 4])
df[, 5] <- sprintf("%.1e", df[, 5])
colnames(df) <- c(
  "Package", "Estimate", "Minimum(rd)", "Maximum(rd)", "Mean(|rd|)"
)
kable(df, align = "llrrr", digits = 20) 
```

The processing times can be compared by the following table (notice the speed 
gain relative to `MASS::rlm()`, which has to estimate the coefficients and SEs 
for 1,009 fixed effects).

```{r TabTiming, echo=FALSE}
df <- bmark_timing[c(3, 4, 6, 2), ]
colnames(df) <- c(
  "Package", "Mean [secs]", "Low CI [secs]", "High CI [secs]"
)
rownames(df) <- NULL
kable(df, align = "lrrr", digits = 3) 
```


## Not yet implemented

- Alternative loss functions
- Weighted regressions


