---
title: "ferols: A Fixed-Effects Robust M Estimator with Huber Loss"
format: gfm
---

**ferols** provides an experimental implementation for a fixed-effects linear 
regression estimator using **Huber M-estimation** with 
**iteratively reweighted least squares (IRLS)**. It is inspired by 
the [package `robtwfe` by David Veenman](https://github.com/dveenman/robtwfe). 
It is build on and designed to integrate tightly with the
[`fixest`](https://lrberge.github.io/fixest/) ecosystem.

It builds on our recent work

> Joachim Gassen and David Veenman (2026): Estimation Precision and 
> Robust Inference in Archival Research, SSRN Working Paper,
> http://dx.doi.org/10.2139/ssrn.4975569.

and combines robust M-estimation (Huber loss) with:

- high-dimensional fixed-effect absorption,
- fast estimation via `fixest::feols`,
- several algorithms to estimate scale in the first step,
- and a `fixest` style variance–covariance interface allowing for
  one-way clustered Huber sandwich standard errors.


## Why?

Standard fixed-effects estimators are sensitive to outliers and heavy-tailed
error distributions. While `fixest` provides fast and reliable estimation
for large linear models with fixed effects, it does not currently offer
robust estimators.

The `ferols` package is a first step to fill this gap. It is intended to be 
useful for settings with high-dimensional fixed effects (e.g. panel data with 
unit and time effects) and offers robust inference via Huber ψ/φ sandwich 
variance estimators, including clustered standard errors.


## Disclaimer 

`ferols` is not meant to replace 
[established R packages for robust regression](https://cran.r-project.org/web/views/Robust.html)
as these provide much more flexible, tested, and rigorous implementations. 
Rather it aims to fill a narrow gap as these packages currently do not allow
to absorb fixed effects during estimation and might thus run in 
performance/convergence problems in setting with many fixed effects.

This package is under active development.  
The API, implementation details, and numerical behavior may and probably will
change. Use with care and 
**do not rely on it to generate reproducible empirical results yet**.


## Installation (development version)

`ferols` is not on CRAN. To install the current development version from GitHub:

```{r, eval=FALSE}
install.packages("remotes")
remotes::install_github("joachim-gassen/ferols")
```

## Function to generate test panel data

For convenience, the package ships with the function `generate_panel_data()`. 
It generates panel data with a composite  error term that has high kurtosis 
by default. The modeled data generating process is as follows

$$
y_{it} = \alpha_i + \lambda_t + x_{it}\beta + z_{it}\gamma + e_{it}
$$

where $\alpha_i$ and $\lambda_t$ are unit and time fixed effects, 
respectively. The composite error is constructed as

$$
e_{it} = u_{it} + r_{it} + \eta_{it}
$$

with three components:

- **Unit component** ($u_{it}$): follows an AR(1) process over time within each 
  unit (serial correlation within units)

- **Time component** ($r_{it}$): follows an AR(1) process over units within each
  time period (cross-sectional dependence within time)

- **Idiosyncratic component** ($eta_{it}$): i.i.d. noise

This construction induces two-way dependence consistent with the simulation 
logic in Petersen (2009).

To generate heavy-tailed errors, a share $p$ of observations is randomly 
selected. The composite errors $e_{it}$ of these observations are multiplied 
by a factor $m$. This produces a mixture error distribution with outliers.


## Example code

```{r exp1, fig.align='center'}
library(ferols)
library(moments)
df <- generate_panel_data(seed = 42, e_mult = 10)
cat(sprintf("Kourtosis of composite error: %.2f", kurtosis(df$e)))
plot(
  df$x, df$y, xlab = "x", ylab = "y", 
  pch = 16, cex = 0.4, col = rgb(0, 0, 0, 0.2)
)
ferols(y ~ x + z | i + t, data = df, vcov = ~ i)
# For comparison: OLS has higher standard errors
fixest::feols(y ~ x + z | i + t, data = df, vcov = ~ i)
```

## A comparison of the precision of OLS and robust M by increasing kurtosis of e

```{r exp2, fig.align="center"}
library(ferols)
library(fixest)
library(moments)

multiplers <- 1:10
res <- matrix(NA_real_, nrow = length(multiplers), ncol =  4)
colnames(res) <- c("m", "k_e", "se_x_feols", "se_x_ferols")
res[,1] <- multiplers
for (r in 1:nrow(res)) {
  df <- generate_panel_data(seed = 1234, e_mult = res[r, "m"])
  res[r, "k_e"] <- kurtosis(df$e)
  fit_ols <- feols(y ~ x + z | i + t, vcov = ~i, data = df)
  res[r, "se_x_feols"] <- fit_ols$se[1]
  fit_rols <- ferols(y ~ x + z | i + t, vcov = ~i, data = df)
  res[r, "se_x_ferols"] <- fit_rols$se[1]
} 
plot(
  main = "Standard error of x coefficient by kurtosis of composite error term",
  res[, "k_e"], res[, "se_x_feols"], type = "l", 
  xlab = bquote("kurtosis("~ e[i*t] ~ ")"), ylab = expression(SE(hat(beta))), 
  lwd  = 2, 
)
lines(
  res[, "k_e"], res[, "se_x_ferols"],
  lwd = 2, lty = 2, col = "blue"
)
points(res[, "k_e"], res[, "se_x_feols"], pch = 16)
points(res[, "k_e"], res[, "se_x_ferols"], col = "blue", pch = 16)
legend(
  "topleft",
  legend = c("OLS (feols)", "Robust M Huber (ferols)"),
  lty = c(1, 2), col = c("black", "blue"), lwd = 2, bty = "n"
)
```

## Not yet implemented (sorted by priority)

- Multi-way clustering
- Alternative loss functions


